---
output:
  html_document:
    keep_md: yes
  pdf_document:
    keep_tex: yes
---
```{r global_options, include = FALSE}
knitr::opts_chunk$set(fig.width = 12, 
                      fig.height = 8, 
                      fig.path = 'figure/',
                      echo = TRUE, 
                      warning = FALSE, 
                      message = FALSE)
```
#Make Effective Data Visualization
###Project: data


###Data
Airport delay statistics (Bureau of Transportation Statistics):
Dataset obtained from: [Dataset link](http://www.transtats.bts.gov/OT_Delay/OT_DelayCause1.asp)

Airport information:

* carrier: Airline code.
* carrier_name: Airline name.
* airport: Airport code.
* airport_name: Airport name.

Airport statistics:

* arr_flights: Number of flights which arrived at the airport.
* arr_del15: Number of flights delayed (>= 15minutes late).
* carrier_ct: Number of flights delayed due to air carrier (e.g. maintenance or crew problems, aircraft cleaning, baggage loading, fueling, etc.).
* weather_ct: Number of flights delayed due to weather.
* nas_ct: Number of flights delayed due to National Aviation System (e.g. non-extreme weather conditions, airport operations, heavy traffic volume, and air traffic control).
* security_ct: Number of flights delayed due to security (e.g. evacuation of a terminal or concourse, re-boarding of aircraft because of security breach, inoperative screening equipment and/or long lines in excess of 29 minutes at screening areas).
* late_aircraft_ct: Number of flights delayed due to a previous flight using the same aircraft being late.
* arr_cancelled: Number of cancelled flights.
* arr_diverted: Number of diverted flights.
* arr_delay: Total time (minutes) of delayed flights.
* carrier_delay: Total time (minutes) of delayed flights due to air carrier.
* weather_delay: Total time (minutes) of delayed flights due to weather.
* nas_delay: Total time (minutes) of delayed flights due to National Aviation System.
* security_delay: Total time (minutes) of delayed flights due to security.
* late_aircraft_delay: Total time (minutes) of delayed flights due to a previous flight using the same aircraft being late.

Airport Database (OpenFlights):
Dataset obtained from: [Dataset link](http://openflights.org/data.html)

Airport Database:

* Airport ID: Unique OpenFlights identifier for this airport.
* Name: Name of airport. May or may not contain the City name.
* City: Main city served by airport. May be spelled differently from Name.
* Country: Country or territory where airport is located.
* airport/FAA: 3-letter FAA code, for airports located in Country "United States of America". 3-letter airport code, for all other airports. Blank if not assigned.
* ICAO: 4-letter ICAO code.
* Blank: if not assigned.
* Latitude: Decimal degrees, usually to six significant digits. Negative is South, positive is North.
* Longitude: Decimal degrees, usually to six significant digits. Negative is West, positive is East.
* Altitude: In feet.
* Timezone: Hours offset from UTC. Fractional hours are expressed as decimals, eg. India is 5.5.
* DST: Daylight savings time. One of E (Europe), A (US/Canada), S (South America), O (Australia), Z (New Zealand), N (None) or U (Unknown). See also: Help: Time
* Tz database time zone: Timezone in "tz" (Olson) format, eg. "America/Los_Angeles". 

Routes Database (OpenFlights):
Dataset obtained from: [Dataset link](http://openflights.org/data.html)

* Airline: 2-letter (airport) or 3-letter (ICAO) code of the airline.
* Airline ID: Unique OpenFlights identifier for airline (see Airline).
* Source airport: 3-letter (airport) or 4-letter (ICAO) code of the source airport.
* Source airport ID: Unique OpenFlights identifier for source airport (see Airport)
* Destination airport: 3-letter (airport) or 4-letter (ICAO) code of the destination airport.
* Destination airport ID: Unique OpenFlights identifier for destination airport (see Airport)
* Codeshare: "Y" if this flight is a codeshare (that is, not operated by Airline, but another carrier), empty otherwise.
* Stops: Number of stops on this flight ("0" for direct)
* Equipment: 3-letter codes for plane type(s) generally used on this flight, separated by spaces.


###1. Loading Packages/ Data
```{r, echo = FALSE}
for (package in c('plyr')) {
  
  if (!require(package, character.only = TRUE, quietly = FALSE)) {
    install.packages(package)
    library(package, character.only = TRUE)
  }
  
}

rm(package)
```

Load the full Bureau of Transportation Statistics dataset.
```{r}
df.raw <- read.csv("data/415580818_112014_2820_airline_delay_causes.csv", header = TRUE, na.strings = c("NA", ""), stringsAsFactors = FALSE)
df_airdelays <- df.raw
df_airdelays[is.na(df_airdelays)] <- 0
```

Load the airport code, latitude and longitude variables from the OpenFlights Airport Database.
```{r}
df.raw <- read.csv("data/airports.dat", header = FALSE, na.strings = c("NA", "NaN", "\\N", ""), stringsAsFactors = FALSE)
#df_airports <- df.raw
df_airports <- df.raw[df.raw[, 4] == "United States", ]
df_airports[is.na(df_airports)] <- 0
df_airports <- df_airports[, c(2, 5, 7, 8)]
colnames(df_airports) <- c("airport_name", "airport", "latitude", "longitude")
```

Load the airport route origin and destination codes from theOpenFlights Route Database.
```{r}
df.raw <- read.csv("data/routes.dat", header = FALSE, na.strings = c("NA", "NaN", "\\N", ""), stringsAsFactors = FALSE)
df_airroutes <- df.raw
df_airroutes[is.na(df_airroutes)] <- 0
df_airroutes <- df_airroutes[, c(3, 5)]
colnames(df_airroutes) <- c("origin", "destination")
df_airroutes <- unique(df_airroutes[, c("origin", "destination")])
```

###2. Pre-process the Data
Drop LGA New York, NY: LaGuardia and EWR Newark, NJ: Newark Liberty International
```{r}
df_airdelays <- df_airdelays[df_airdelays[, "airport"]!= "LGA",]
df_airdelays <- df_airdelays[df_airdelays[, "airport"]!= "EWR",]
```

Get state/region from airport names:
```{r}
state <- c("CO", "NV", "CA", "OR", "AZ", "WA", "UT",
  "MI", "IL", "MN",
  "MA", "NY", "PA",
  "GA", "MD", "NC", "DC", "TX", "FL", "GA")

region <- c("West", "West", "West", "West", "West", "West", "West",
  "Mid West", "Mid West", "Mid West",
  "North East", "North East", "North East",
  "South", "South", "South", "South", "South", "South", "South")

df_stateref <- data.frame(state = state, region = region, stringsAsFactors = FALSE)

df_airdelays$state = NA
df_airdelays$region = NA

for (i in 1:nrow(df_airdelays)) {
  
  temp <- df_airdelays[i, "airport_name"]
  temp <- unlist(strsplit(temp, '[,]'))[2]
  temp <- unlist(strsplit(temp, '[:]'))[1]
  temp <- gsub(" ", "", temp, fixed = TRUE)
  df_airdelays[i, "state"] <- temp
 
  temp = df_stateref[df_stateref[, "state"] == temp, 2][1]
  df_airdelays[i, "region"] <- temp
   
}

rm(state, region, temp)
```

Ensure that airport codes match between each dataset. Drop any entries which do not appear in all three datasets.
```{r}
#Ensure common index between datasets (better way to do this?):
for (i in 1:2) {

  df_airdelays <- df_airdelays[(df_airdelays$airport %in% df_airports$airport), ]
  df_airdelays <- df_airdelays[(df_airdelays$airport %in% df_airroutes$origin), ]
  df_airdelays <- df_airdelays[(df_airdelays$airport %in% df_airroutes$destination), ]
  
  df_airports <- df_airports[(df_airports$airport %in% df_airroutes$origin), ]
  df_airports <- df_airports[(df_airports$airport %in% df_airroutes$destination), ]
  df_airports <- df_airports[(df_airports$airport %in% df_airdelays$airport), ]
  
  df_airroutes <- df_airroutes[(df_airroutes$origin %in% df_airports$airport), ]
  df_airroutes <- df_airroutes[(df_airroutes$origin %in% df_airdelays$airport), ]
  df_airroutes <- df_airroutes[(df_airroutes$destination %in% df_airports$airport), ]
  df_airroutes <- df_airroutes[(df_airroutes$destination %in% df_airdelays$airport), ]

}
```

Create annual airport flight delay summary metrics
```{r}
df_annairdelays <- ddply(df_airdelays, .(year, airport, airport_name, state, region), summarize,
  arr_flights = sum(arr_flights),
  arr_del15 = sum(arr_del15),
  carrier_ct = sum(carrier_ct),
  weather_ct = sum(X.weather_ct),
  nas_ct = sum(nas_ct),
  security_ct = sum(security_ct),
  late_aircraft_ct = sum(late_aircraft_ct))
```

Create annual region flight delay summary metrics
```{r}
df_annregdelays <- ddply(df_airdelays, .(year, region), summarize,
  arr_flights = sum(arr_flights),
  arr_del15 = sum(arr_del15))
```

Create airports dataset
```{r}
temp1 <- data.frame(year = integer(),
   airport = character(),
   airport_name = character(),
   state = character(),
   region = character(),
   latitude = numeric(),
   longitude = numeric(),
   delay_type = character(),
   delay_prop = numeric(), 
   stringsAsFactors = FALSE)

df_lbl <- c("Air Carrier", "Weather", "NAS", "Security", "Previous Flight")
df_col <- c("carrier_ct", "weather_ct", "nas_ct", "security_ct", "late_aircraft_ct")


for (i in 1:length(df_lbl)) {
  
  temp2 <- df_annairdelays
  temp2$delay_type <- df_lbl[i]
  temp2$delay_prop <- df_annairdelays[[df_col[i]]] / df_annairdelays$arr_flights
  
  for (i in 1:nrow(temp2)) {
    
    airport <- temp2[i, "airport"]
    temp2[i, "latitude"] <- df_airports[which(df_airports$airport == airport), 3]
    temp2[i, "longitude"] <- df_airports[which(df_airports$airport == airport), 4]
    
  }
  
  temp2 <- temp2[, c(1, 2, 3, 4, 5, 15, 16, 13, 14)]
  
  temp1 <- rbind.data.frame(temp1, temp2)
  
}


df_airports <- temp1

rm(temp1, temp2, df_lbl, df_col, airport)
```

Create region dataset
```{r}
temp <- data.frame(year = integer(),
   state = character(),
   region = character(),
   delay_prop = numeric())

temp <- df_annregdelays
temp$delay_prop <- df_annregdelays$arr_del15 / df_annregdelays$arr_flights

df_airregion <- temp

rm(temp)
```

Create airport sentence data
```{r}
df_airsentence <- data.frame(airport = character(),
   delay_curr = numeric(),
   delay_trend = character(),
   delay_likely = character(),
   delay_comp = character(), 
   stringsAsFactors = FALSE)
  

airports <- unique(df_airports$airport)

for (i in 1:length(airports)) {
  
  airport <- airports[i]
  
  df_airsentence[i, 1] = airport
  
  temp1 <- ddply(df_airports, .(year, airport), summarize,
    delay_prop = sum(delay_prop))
  
  temp1 <- temp1[temp1[, "airport"] == airport, ]

  delay_curr <- temp1[temp1[, "year"] == 2015, 3]
  
  df_airsentence[i, 2] = delay_curr
  
  lm <- lm(temp1$delay_prop ~ temp1$year)
  
  delay_trend <- if (summary(lm)$coefficients[2, 1] < -0.001) {
    "decreased"
  } else if (summary(lm)$coefficients[2, 1] >= -0.001 & summary(lm)$coefficients[2, 1] <= 0.001) {
    "generally remained steady"
  } else {
    "increased"
  }
  
  df_airsentence[i, 2] = round(df_airsentence[i, 2], digits = 2) * 100
  
  df_airsentence[i, 3] = delay_trend
  
  temp2 <- ddply(df_airports, .(airport, region, delay_type), summarize,
    delay_prop = mean(delay_prop))
  
  temp2 <- temp2[temp2[, "airport"] == airport, ]
  
  delay_likely <- temp2[which.max( temp2[, 4] ), 3]
  
  df_airsentence[i, 4] = delay_likely
  
  temp3 <- ddply(df_airregion, .(region), summarize,
    delay_prop = mean(delay_prop))
  
  delay_comp <- if (sum(temp2[, 4]) < temp3[temp3[, "region"] == temp2[1, 2], 2]) {
    "better"
  } else {
    "worse"
  }
  
  df_airsentence[i, 5] = delay_comp
  
}

rm(temp1, temp2, temp3, airport, airports, delay_curr, delay_trend, delay_likely, delay_comp, lm)

```

Add airport sentence data to airports dataset
```{r}
df_airports$delay_curr <- 0
df_airports$delay_trend <- ""
df_airports$delay_likely <- ""
df_airports$delay_comp <- ""

for (i in 1:nrow(df_airports)) {
  
  airport <- df_airports[i, 2]
  df_airports[i, 10] <- df_airsentence[which(df_airsentence$airport == airport), 2]
  df_airports[i, 11] <- df_airsentence[which(df_airsentence$airport == airport), 3]
  df_airports[i, 12] <- df_airsentence[which(df_airsentence$airport == airport), 4]
  df_airports[i, 13] <- df_airsentence[which(df_airsentence$airport == airport), 5]
  
}

rm(airport)
```

Write csv
```{r}
write.csv(df_airroutes, file = "data/airroutes.csv")
write.csv(df_airports, file = "data/airports.csv")
write.csv(df_airregion, file = "data/airregion.csv")
```