---
output:
  html_document:
    keep_md: yes
  pdf_document:
    keep_tex: yes
---
```{r global_options, include = FALSE}
knitr::opts_chunk$set(fig.width = 12, 
                      fig.height = 8, 
                      fig.path = 'figure/',
                      echo = TRUE, 
                      warning = FALSE, 
                      message = FALSE)
```
#Make Effective Data Visualization
###Project: ND_Project6


###Introduction
This repository holds results for the Udacity Nanodegree Project: Make Effective Data Visualizations. For this project, I selected the 'flights' dataset and exercised d3.js in order to make an interactive map of the USA which allows visualization of airport statistics.

The interactive map was based on the d3.js example found here: [d3.js Symbol Map](http://mbostock.github.io/d3/talk/20111116/airports.html)


###Data
The flights dataset provided by Udacity originally came from the Bureau of Transportation Statistics. It contains 15 statistics on airports located within the USA, arranged by time and airport.

The dataset was obtained from: [Dataset link](http://www.transtats.bts.gov/OT_Delay/OT_DelayCause1.asp)

Airport information:

* carrier: Airline code.
* carrier_name: Airline name.
* airport: Airport code.
* airport_name: Airport name.

Airport statistics:

* arr_flights: Number of flights which arrived at the airport.
* arr_del15: Number of flights delayed (>= 15minutes late).
* carrier_ct: Number of flights delayed due to air carrier (e.g. maintenance or crew problems, aircraft cleaning, baggage loading, fueling, etc.).
* weather_ct: Number of flights delayed due to weather.
* nas_ct: Number of flights delayed due to National Aviation System (e.g. non-extreme weather conditions, airport operations, heavy traffic volume, and air traffic control).
* security_ct: Number of flights delayed due to security (e.g. evacuation of a terminal or concourse, re-boarding of aircraft because of security breach, inoperative screening equipment and/or long lines in excess of 29 minutes at screening areas).
* late_aircraft_ct: Number of flights delayed due to a previous flight using the same aircraft being late.
* arr_cancelled: Number of cancelled flights.
* arr_diverted: Number of diverted flights.
* arr_delay: Total time (minutes) of delayed flights.
* carrier_delay: Total time (minutes) of delayed flights due to air carrier.
* weather_delay: Total time (minutes) of delayed flights due to weather.
* nas_delay: Total time (minutes) of delayed flights due to National Aviation System.
* security_delay: Total time (minutes) of delayed flights due to security.
* late_aircraft_delay: Total time (minutes) of delayed flights due to a previous flight using the same aircraft being late.


In addition to the data described above, additional data was downloaded from OpenFlights. This included the OpenFlights Airport Database (airports.dat) and the OpenFlights Routes Database (routes.dat).

These datasets were obtained from: [Dataset link](http://openflights.org/data.html)

Airport Database:

* Airport ID: Unique OpenFlights identifier for this airport.
* Name: Name of airport. May or may not contain the City name.
* City: Main city served by airport. May be spelled differently from Name.
* Country: Country or territory where airport is located.
* airport/FAA: 3-letter FAA code, for airports located in Country "United States of America". 3-letter airport code, for all other airports. Blank if not assigned.
* ICAO: 4-letter ICAO code.
* Blank: if not assigned.
* Latitude: Decimal degrees, usually to six significant digits. Negative is South, positive is North.
* Longitude: Decimal degrees, usually to six significant digits. Negative is West, positive is East.
* Altitude: In feet.
* Timezone: Hours offset from UTC. Fractional hours are expressed as decimals, eg. India is 5.5.
* DST: Daylight savings time. One of E (Europe), A (US/Canada), S (South America), O (Australia), Z (New Zealand), N (None) or U (Unknown). See also: Help: Time
* Tz database time zone: Timezone in "tz" (Olson) format, eg. "America/Los_Angeles". 

Routes Database:

* Airline: 2-letter (airport) or 3-letter (ICAO) code of the airline.
* Airline ID: Unique OpenFlights identifier for airline (see Airline).
* Source airport: 3-letter (airport) or 4-letter (ICAO) code of the source airport.
* Source airport ID: Unique OpenFlights identifier for source airport (see Airport)
* Destination airport: 3-letter (airport) or 4-letter (ICAO) code of the destination airport.
* Destination airport ID: Unique OpenFlights identifier for destination airport (see Airport)
* Codeshare: "Y" if this flight is a codeshare (that is, not operated by Airline, but another carrier), empty otherwise.
* Stops: Number of stops on this flight ("0" for direct)
* Equipment: 3-letter codes for plane type(s) generally used on this flight, separated by spaces.


###1. Loading Packages/ Data
```{r, echo = FALSE}
for (package in c('plyr')) {
  
  if (!require(package, character.only = TRUE, quietly = FALSE)) {
    install.packages(package)
    library(package, character.only = TRUE)
  }
  
}
```

Load the full Bureau of Transportation Statistics dataset.
```{r}
df.raw <- read.csv("data/415580818_112014_2820_airline_delay_causes.csv", header = TRUE, na.strings = c("NA", ""), stringsAsFactors = FALSE)
df_airdelays <- df.raw
df_airdelays[is.na(df_airdelays)] <- 0
```

Load the airport code, latitude and longitude variables from the OpenFlights Airport Database.
```{r}
df.raw <- read.csv("data/airports.dat", header = FALSE, na.strings = c("NA", "NaN", "\\N", ""), stringsAsFactors = FALSE)
#df_airports <- df.raw
df_airports <- df.raw[df.raw[, 4] == "United States", ]
df_airports[is.na(df_airports)] <- 0
df_airports <- df_airports[, c(2, 5, 7, 8)]
colnames(df_airports) <- c("airport_name", "airport", "latitude", "longitude")
```

Load the airport route origin and destination codes from theOpenFlights Route Database.
```{r}
df.raw <- read.csv("data/routes.dat", header = FALSE, na.strings = c("NA", "NaN", "\\N", ""), stringsAsFactors = FALSE)
df_airroutes <- df.raw
df_airroutes[is.na(df_airroutes)] <- 0
df_airroutes <- df_airroutes[, c(3, 5)]
colnames(df_airroutes) <- c("origin", "destination")
df_airroutes <- unique(df_airroutes[, c("origin", "destination")])
```

###2. Pre-process the Data
Ensure that airport codes match between each dataset. Drop any entries which do not appear in all three datasets.
```{r}
#Ensure common index between datasets (better way to do this?):
for (i in 1:2) {

  df_airdelays <- df_airdelays[(df_airdelays$airport %in% df_airports$airport), ]
  df_airdelays <- df_airdelays[(df_airdelays$airport %in% df_airroutes$origin), ]
  df_airdelays <- df_airdelays[(df_airdelays$airport %in% df_airroutes$destination), ]
  
  df_airports <- df_airports[(df_airports$airport %in% df_airroutes$origin), ]
  df_airports <- df_airports[(df_airports$airport %in% df_airroutes$destination), ]
  df_airports <- df_airports[(df_airports$airport %in% df_airdelays$airport), ]
  
  df_airroutes <- df_airroutes[(df_airroutes$origin %in% df_airports$airport), ]
  df_airroutes <- df_airroutes[(df_airroutes$origin %in% df_airdelays$airport), ]
  df_airroutes <- df_airroutes[(df_airroutes$destination %in% df_airports$airport), ]
  df_airroutes <- df_airroutes[(df_airroutes$destination %in% df_airdelays$airport), ]

}
```

Create flight delay summary metrics
```{r}
df_sumairdelays <- ddply(df_airdelays, .(airport), summarize,
  arr_flights = sum(arr_flights),
  arr_del15 = sum(arr_del15),
  carrier_ct = sum(carrier_ct),
  weather_ct = sum(X.weather_ct),
  nas_ct = sum(nas_ct),
  security_ct = sum(security_ct),
  late_aircraft_ct = sum(late_aircraft_ct))
```

Create annual flight delay summary metrics
```{r}
df_annairdelays <- ddply(df_airdelays, .(year, airport, airport_name), summarize,
  arr_flights = sum(arr_flights),
  arr_del15 = sum(arr_del15),
  carrier_ct = sum(carrier_ct),
  weather_ct = sum(X.weather_ct),
  nas_ct = sum(nas_ct),
  security_ct = sum(security_ct),
  late_aircraft_ct = sum(late_aircraft_ct))
```

Create airports dataset
```{r}
temp1 <- data.frame(year = numeric(),
   airport = character(),
   airport_name = character(),
   latitude = numeric(),
   longitude = numeric(),
   delay_type = numeric(),
   delay_prop = numeric())

df_lbl <- c("Air Carrier", "Weather", "NAS", "Security", "Previous Flight")
df_col <- c("carrier_ct", "weather_ct", "nas_ct", "security_ct", "late_aircraft_ct")


for (i in 1:length(df_lbl)) {
  
  temp2 <- df_annairdelays
  temp2$delay_type <- df_lbl[i]
  temp2$delay_prop <- df_annairdelays[[df_col[i]]] / df_annairdelays$arr_flights
  
  for (i in 1:nrow(temp2)) {
    
    airport <- temp2[i, "airport"]
    temp2[i, "latitude"] <- df_airports[which(df_airports$airport == airport), 3]
    temp2[i, "longitude"] <- df_airports[which(df_airports$airport == airport), 4]
    
  }
  
  temp2 <- temp2[, c(1, 2, 3, 13, 14, 11, 12)]
  
  temp1 <- rbind(as.matrix(temp1), as.matrix(temp2))
  
}

df_airports <- temp1
```

Write csv
```{r}
write.csv(df_airroutes, file = "data/airroutes.csv")
write.csv(df_airports, file = "data/airports.csv")
```
